---
title: "Analysis of COVID-19 cases and deaths in the U.S."
date: "2023-02-24"
output: 
  pdf_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction  
## Data source  
The data used in this analysis is provided by Johns Hopkins University Center
for Systems Science and Engineering (JHU CSSE). It is connected with an article 
in The Lancet titled "An interactive web-based dashboard to track COVID-19 in 
real time" which can be found at: https://www.thelancet.com/journals/laninf/article/PIIS1473-3099(20)30120-1/fulltext.  
The data repository is maintained for the 2019 Novel Coronavirus Visual Dashboard
that is maintained by JHU CSSE.  The data files can be found at:
https://github.com/CSSEGISandData/COVID-19.
The data sets specifically used here come from the time series folder in the 
repository and include both the U.S. confirmed case and death counts.

## Retrieving the data  
```{r, get_the_data}
library(tidyverse)
url_root <- "https://github.com/CSSEGISandData/COVID-19/raw/master/csse_covid_19_data/csse_covid_19_time_series/"
filenames <- c("time_series_covid19_confirmed_US.csv", "time_series_covid19_deaths_US.csv")
urls <- str_c(url_root, filenames)
init_us_cases <- read_csv(urls[1])
init_us_deaths <- read_csv(urls[2])

```

## Data description  
The confirmed cases and COVID deaths data sets include daily listings starting
in January 2020 and detailed per U.S. state and county.
```{r, quick_view}
head(init_us_cases, 10)
head(init_us_deaths, 10)
```

## Goals of the analysis  
Goals of this analysis include:  
1.  Visualize the overall trend of confirmed cases and deaths  
2.  Zoom in on a specific state to see its trend  
3.  Determine the U.S. states with the highest and lowest rates  
4.  Model the relationship between cases and deaths  
5.  Visualize the total cases across the U.S. geographic regions and divisions  
6.  Model the relationship between case rate and political bias  
7.  Model the relationship between case rate and state GDP  

## Pivoting and joining the data  
The data is provided in "wide" format and is converted to "long" format here.
Additionally, unnecessary columns are removed.  The confirmed cases and number
of deaths combined into a single data frame so they can be analyzed together.
```{r, pivot_and_combine}
library(lubridate)
us_cases <- init_us_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

us_deaths <- init_us_deaths %>%
  pivot_longer(cols = -(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))
us <- us_cases %>%
  full_join(us_deaths)

```

# Initial Analysis and Visualization  
## Viewing U.S. totals  
```{r, US_by_state}
US_by_state <- us %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mil = deaths * 1000000 / Population) %>%
  select(Province_State, Country_Region, date,
         cases, deaths, deaths_per_mil, Population) %>%
  ungroup()
```
```{r, US_totals}
US_totals <- US_by_state %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mil = deaths * 1000000 / Population) %>%
  select(Country_Region, date,
         cases, deaths, deaths_per_mil, Population) %>%
  ungroup()

US_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID-19 in the U.S. during the pandemic", y=NULL)
```

## Determining how infection spread over time  
## Building out data by U.S. state  
```{r, eval_US}
 US_by_state <- US_by_state %>%
   mutate(new_cases = cases - lag(cases),
          new_deaths = deaths - lag(deaths))
US_totals <- US_totals %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))
tail(US_totals %>% select(new_cases, new_deaths, everything()))
US_totals %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID-19 new cases and new deaths in the U.S.", y=NULL)

state <- "New York"
US_by_state %>%
  filter(Province_State == state) %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID-19 in ", state), y=NULL)

```

## Normalizing cases and deaths by population
```{r, totals_by_state}
US_state_totals <- US_by_state %>%
  group_by(Province_State) %>%
  summarize(deaths = max(deaths), cases = max(cases),
            population = max(Population),
            cases_per_thou = 1000 * cases / population,
            deaths_per_thou = 1000 * deaths / population) %>%
  filter(cases > 0, population > 0)
```
The following is a view of the 10 lowest states and 10 highest states as
ranked by number of deaths per thousand.
```{r view_per_thousand}
US_state_totals %>% slice_min(deaths_per_thou, n=10) %>%
  select(deaths_per_thou, cases_per_thou, everything())

US_state_totals %>% slice_max(deaths_per_thou, n=10) %>%
  select(deaths_per_thou, cases_per_thou, everything())
```
## Predictive model for relationship between cases and deaths  
The following visual shows the relationship between the number of deaths per
thousand and cases per thousand.
```{r some_modeling}
mod <- lm(deaths_per_thou ~ cases_per_thou, data = US_state_totals)
summary(mod)
US_tot_w_pred <- US_state_totals %>% mutate(pred = predict(mod))
US_tot_w_pred %>% ggplot() +
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red") +
  labs(title = "Predictive relationship between cases and deaths",
       x = "Cases per Thousand", y = "Deaths per Thousand")
```

# U.S. States by Region and Division
The U.S. Census Bureau divides the U.S. geographically into four regions and
nine divisions.  Here, the data is grouped and summarized into the cases and
deaths per thousand per geographic region and division.
```{r, regions_and_divisions}
# Merge in the region / division data
US_states_by_region_div <- read_csv("US_regions_and_divisions.csv")
US_by_state <- US_by_state %>% rename("State" = "Province_State")
US_by_state <- merge(US_by_state, US_states_by_region_div, by="State")

# Combine the region and division columns
US_by_state <- US_by_state %>%
  #mutate(region = get_US_region[Province_State]) %>%
  #mutate(division = get_US_division[Province_State]) %>%
  #mutate(ABV = get_abv[Province_State]) %>%
  unite("Region_Div",
        Region:Division,
        sep = ": ",
        na.rm = TRUE,
        remove = FALSE)

US_by_state$Region_Div[US_by_state$Region_Div == ""] <- "Other"

US_regionDiv_totals <- US_by_state %>%
  group_by(Region_Div) %>%
  summarize(deaths = max(deaths), cases = max(cases),
            population = max(Population),
            cases_per_thou = 1000 * cases / population,
            deaths_per_thou = 1000 * deaths / population) %>%
  filter(cases > 0, population > 0)

US_regionDiv_totals %>%
  ggplot(aes(x = Region_Div, y = cases_per_thou, fill = Region_Div)) +
    geom_bar(stat = "identity") +
    labs(title = "COVID total cases per thousand by U.S. region and dvision", 
         x = NULL, y = "Cases per thousand") + 
    theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(),
          legend.key.height = unit(0.5, 'cm'),
          legend.key.width = unit(0.5, 'cm'))

```
The chart above shows that, when normalized by population, there was not a 
significant amount of variation across geographic regions and divisions.
# Using 2020 Presidential Election results  
The results of the 2020 presidential election are available through many sources
online.  One example is here, https://www.cookpolitical.com/2020-national-popular-vote-tracker.
The following examines the relationship between case and deaths rates and the
popular vote results.  Specifically, it looks at how states fared that voted
more or less Democratic (voted for Joe Biden).
```{r, pres_election}
US_state_totals_w_geopol <- US_state_totals %>% rename("State" = "Province_State")
US_state_totals_w_geopol <- merge(US_state_totals_w_geopol, US_states_by_region_div, by="State")
US_state_totals_w_geopol <- US_state_totals_w_geopol %>%
  unite("Region_Div",
        Region:Division,
        sep = ": ",
        na.rm = TRUE,
        remove = FALSE)

election_results <- read_csv("2020presgeresults.csv")
election_results <- election_results %>% drop_na()
US_state_totals_w_geopol <- merge(US_state_totals_w_geopol, election_results, by="ABV")

US_state_totals_w_geopol <- US_state_totals_w_geopol %>%
  mutate(DemVotes = BIDEN/TotalVotes) %>%
  drop_na()

US_state_totals_w_geopol %>%
  ggplot(aes(x = DemVotes, y = cases_per_thou)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths_per_thou, color = "deaths")) +
  geom_point(aes(y = deaths_per_thou, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "Cases & Deaths according to how Democratic the vote", 
       x = "Number of Voters who voted for Democrat (Biden)", y=NULL)

```
# Model of Election Data  
```{r, case_and_death_model}
case_model <- lm(cases_per_thou ~ DemVotes, data = US_state_totals_w_geopol)
summary(case_model)
case_pred <- US_state_totals_w_geopol %>% mutate(pred = predict(case_model))
case_pred %>% ggplot() +
  geom_point(aes(x = DemVotes, y = cases_per_thou), color = "blue") +
  geom_point(aes(x = DemVotes, y = pred), color = "red") +
  scale_x_continuous(labels = scales::percent) +
  labs(title = "Relationship between 2020 election party bias and COVID cases", 
       x = "Number of Voters who voted for Democrat (Biden)", y = "Cases per thousand") + 
  theme(legend.position = "bottom")

death_model <- lm(deaths_per_thou ~ DemVotes, data = US_state_totals_w_geopol)
summary(death_model)
death_pred <- US_state_totals_w_geopol %>% mutate(pred = predict(death_model))
death_pred %>% ggplot() +
  geom_point(aes(x = DemVotes, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = DemVotes, y = pred), color = "red") +
  scale_x_continuous(labels = scales::percent) +
  labs(title = "Relationship between 2020 election party bias and COVID deaths", 
       x = "Number of Voters who voted for Democrat (Biden)", y = "Cases per thousand") + 
  theme(legend.position = "bottom")
  

```

# Economic base  

The GDP per state and GDP per capita per state is available in many places online
such as at: https://www.statista.com/statistics/248063/per-capita-us-real-gross-domestic-product-gdp-by-state/.  Here, the analysis looks at the COVID case and death rates in
relation to state GDP.  The top 10 and lowest 10 states are examined to see if
richer or poorer states fared better or worse through the pandemic.
```{r, econ_base}
top_states_by_gdp <- as.factor(
  c("California", "Texas", "New York", "Florida", "Illinois", 
    "Pennsylvania", "Ohio", "Washington", "Georgia", "New Jersey"))
bottom_states_by_gdp <- as.factor(
  c("West Virginia", "Delaware", "Maine", "Rhode Island", "North Dakota",
    "South Dakota", "Montana", "Alaska", "Wyoming", "Vermont"))
top_states_gdp_per_capita <- as.factor(
  c("New York", "Massachusetts", "Washington", "Connecticut", "California",
    "Delaware", "Illinois", "Aalska", "Maryland", "North Dakota"))
bottom_states_gdp_per_capita <- as.factor(
  c("Maine", "Montana", "New Mexico", "Kentucky", "South Carolina", 
    "Idaho", "Alabama", "Arkansas", "West Virginia", "Mississippi"))
top_states_df <- data.frame(GDP_Rank = c(1:10), State = top_states_by_gdp)
bottom_states_df <- data.frame(GDP_Rank = c(41:50), State = bottom_states_by_gdp)
states_by_gdp <- rbind(top_states_df, bottom_states_df)

US_state_totals_w_econ <- US_state_totals %>% rename("State" = "Province_State")
US_state_totals_w_econ <- merge(US_state_totals_w_econ, states_by_gdp, by="State")

US_state_totals_w_econ %>%
  ggplot(aes(x = GDP_Rank, y = cases_per_thou, fill = GDP_Rank)) +
    geom_bar(stat = "identity") +
    labs(title = "COVID total cases per thousand by state's GDP Rank", 
         x = NULL, y = "Cases per thousand") +
     scale_x_continuous(breaks = seq(from = 0, to = 2.5, by = 0.25)) +
    theme(legend.position = "bottom",
          axis.text.x=element_blank(), axis.ticks.x=element_blank())

```

## Predictive model for relationship between GDP and COVID case rate  
```{r, econ_model}
econ_model <- lm(cases_per_thou ~ GDP_Rank, data = US_state_totals_w_econ)
summary(econ_model)
econ_pred <- US_state_totals_w_econ %>% mutate(pred = predict(econ_model))
econ_pred %>% ggplot() +
  geom_point(aes(x = GDP_Rank, y = cases_per_thou), color = "blue") +
  geom_point(aes(x = GDP_Rank, y = pred), color = "red") +
  #scale_x_continuous(breaks = c(1:10, 41:50)) +
  labs(title = "Relationship between state GDP Rank and COVID cases", 
       x = "GDP Rank", y = "Cases per thousand") + 
  theme(legend.position = "bottom")

```
The above chart and model show that there is essentially no relationship
between case and deaths rates with respect to state GDP.

# Conclusion  
## Summary  
This analysis has found that the number of cases of COVID initially skyrocketed 
but has largely flattened out over the past year.  The rates in New York are very
similar to the rest of the country.  There is a loose relationship between the
number of confirmed cases and the death rate.
Looking at the data geographically, the cases per thousand were slightly higher 
in some the South Region's divisions, notably the division containing Kentucky, 
Tennessee, Mississippi, and Alabama.
Using data from the 2020 Presidential election, a weak political bias can be
seen in which states that lean more Democratic and less Republican were likely
to have somewhat lower confirmed cases and deaths.  Finally, the analysis has
shown that a state's GDP is a weak indicator of case rates.  Specifically, lower
GDPs also have lower case rates.

## Sources of Bias  
Sources of bias in COVID reporting data have been heavily discussed in the media.
There is inconsistency across states and hospitals as to when and how COVID is 
reported.  For example, listing a COVID death when there is a co-morbidity has
been a key controversy.  